---
title: "Dynamic changes of DTE isoforms fitting specific patterns"
author: "Madison Mehlferber"
date: "2024-09-16"
output: html_document
---

Load the needed packages 
```{r}
library(here)
library(edgeR)
library(dplyr)
library(tidyverse)
library(ggplot2)
library(knitr)
library(pheatmap)
library(RColorBrewer)
```

Define the project root as the current working directory
```{r}
# Define the project root as the current working directory
project_root <- '/Volumes/sheynkman/projects/Mohi_MDS_LRP/20_visualization/DTE_MM_scripts'

# Define the path for the results folder inside the project root (on the same level as the scripts folder)
results_folder <- file.path(project_root, "results")

# Create the results folder if it doesn't exist
if (!file.exists(results_folder)) {
  dir.create(results_folder)
}

# Define the path for the "01_experiment" subfolder within results
experiment_subfolder <- file.path(results_folder, "10_differential_transcript_expression")
if (!file.exists(experiment_subfolder)) {
  dir.create(experiment_subfolder)
}

# Define the subdirectories for plots and output_table inside "01_experiment"
plots_subfolder <- file.path(experiment_subfolder, "plots")
output_table_subfolder <- file.path(experiment_subfolder, "output_table")

# Create the plots subfolder if it doesn't exist
if (!file.exists(plots_subfolder)) {
  dir.create(plots_subfolder)
}

# Create the output_table subfolder if it doesn't exist
if (!file.exists(output_table_subfolder)) {
  dir.create(output_table_subfolder)
}

# Print paths for debugging
cat("Results folder path:", results_folder, "\n")
cat("Experiment folder path:", experiment_subfolder, "\n")
cat("Plots subfolder path:", plots_subfolder, "\n")
cat("Output table subfolder path:", output_table_subfolder, "\n")
```

Read in and prepare the data
```{r}
counts <- read.csv("/Volumes/sheynkman/projects/Mohi_MDS_LRP/01_isoseq/collapse/merged.collapsed.flnc_count.txt", header = TRUE, row.names = 1)
group <- factor(c("Q157R", "Q157R", "Q157R", "WT", "WT", "WT"))
dge <- DGEList(counts = counts, group = group)
cpm_matrix <- cpm(dge, log = FALSE)
cpm_df <- as.data.frame(cpm_matrix)
cpm_df$transcript_id <- rownames(cpm_df)

cpm_long <- cpm_df %>%
  pivot_longer(cols = -transcript_id, names_to = "sample", values_to = "CPM") %>%
  mutate(group = case_when(
    str_detect(sample, "BioSample_[1-3]") ~ "Q157R",
    str_detect(sample, "BioSample_[4-6]") ~ "WT",
    TRUE ~ NA_character_
  ))

dynamic_transcripts <- read.table(
  file.path(output_table_subfolder, "transcripts_w_dynamic_via_group_comparison_top500.txt"), 
  header = TRUE, sep = "\t")

# Load annotations
gene_map <- read.table("/Volumes/sheynkman/projects/Mohi_MDS_LRP/01_reference_tables/ensg_gene.tsv", header = TRUE, sep = "\t")
colnames(gene_map) <- c("associated_gene", "gene_name")

sqanti <- read.table("/Volumes/sheynkman/projects/Mohi_MDS_LRP/02_sqanti/MDS_classification.txt", sep = "\t", header = TRUE)
sqanti <- sqanti %>%
  mutate(display_name = case_when(
    !is.na(associated_transcript) & associated_transcript != "novel" ~ associated_transcript,
    grepl("^novelGene_", associated_gene) ~ associated_gene,
    TRUE ~ associated_gene
  ))

# Merge metadata
dynamic_transcripts <- dynamic_transcripts %>%
  left_join(gene_map, by = "associated_gene") %>%
  left_join(sqanti[, c("isoform", "display_name")], by = c("isoform" = "isoform"))

dynamic_transcripts$display_name <- gsub("\\.\\d+$", "", dynamic_transcripts$display_name)
gene_map$gene_name <- gsub("\\.\\d+$", "", gene_map$gene_name)

```

Merge with dynamic transcripts
```{r}
dynamic_cpm_long <- cpm_long %>%
  filter(transcript_id %in% dynamic_transcripts$isoform) %>%
  left_join(dynamic_transcripts %>% select(isoform, associated_gene, gene_name, display_name), 
            by = c("transcript_id" = "isoform"))
```
Plot example: per gene
```{r}
# Preview one gene manually
gene_of_interest <- "ENSMUSG00000023025.17"  # Replace with a gene of your choice

# Optionally convert to gene symbol if available
symbol <- gene_map %>%
  filter(associated_gene == gene_of_interest) %>%
  pull(gene_name) %>%
  unique()

plot_title <- if (length(symbol) > 0 && !is.na(symbol) && symbol != "") {
  paste("Isoform Expression for", symbol, "(WT vs Q157R)")
} else {
  paste("Isoform Expression for", gene_of_interest, "(WT vs Q157R)")
}

gene_plot_data <- dynamic_cpm_long %>%
  filter(associated_gene == gene_of_interest)

# Ensure WT is on the left, Q157R on the right
gene_plot_data$group <- factor(gene_plot_data$group, levels = c("WT", "Q157R"))

ggplot(gene_plot_data, aes(x = group, y = CPM, color = group)) +
  geom_boxplot() +
  geom_jitter(width = 0.2, alpha = 0.7) +
  facet_wrap(~ display_name, scales = "free") +
  theme_minimal() +
  labs(title = plot_title,
       x = "Condition",
       y = "CPM") +
  theme(axis.text.x = element_text(angle = 90))
```
Plot isoform expression per gene
```{r}
# Add group for plotting
dynamic_transcripts$group <- ifelse(dynamic_transcripts$logFC > 0, "Q157R", "WT")
dynamic_transcripts$group <- factor(dynamic_transcripts$group, levels = c("WT", "Q157R"))

# Identify genes with multiple dynamic isoforms
non_unique <- dynamic_transcripts %>%
  filter(duplicated(associated_gene) | duplicated(associated_gene, fromLast = TRUE)) %>%
  arrange(associated_gene)

genes_to_plot <- unique(non_unique$associated_gene)

# Loop through genes and generate plots
for (gene in genes_to_plot) {
  gene_data <- dynamic_transcripts %>% filter(associated_gene == gene)
  
  # Ensure group factor levels are set consistently
  gene_data$group <- factor(gene_data$group, levels = c("WT", "Q157R"))
  
  gene_symbol <- unique(gene_data$gene_name)[1]
  
  title_label <- if (!is.na(gene_symbol) && gene_symbol != "") {
    paste("Expression of", gene_symbol, "Isoforms (WT vs Q157R)")
  } else {
    paste("Expression of", gene, "Isoforms (WT vs Q157R)")
  }

  p <- ggplot(gene_data, aes(x = group, y = logCPM, color = group)) +
    geom_boxplot() +
    geom_jitter(width = 0.2, alpha = 0.7) +
    facet_wrap(~ display_name, scales = "free") +
    theme_minimal() +
    labs(title = title_label,
         x = "Condition", 
         y = "logCPM") +
    theme(axis.text.x = element_text(angle = 90))

  filename <- paste0(gsub("[:/\\\\\\s]", "_", ifelse(!is.na(gene_symbol), gene_symbol, gene)), "_isoform_expression_WT_vs_Q157R.pdf")

  ggsave(file.path(plots_subfolder, filename),
         plot = p, width = 6, height = 5)
}
```
Heat map for filtered transcripts - top 500
```{r}
transcript_matrix <- dynamic_cpm_long %>%
  select(transcript_id, sample, CPM) %>%
  pivot_wider(names_from = sample, values_from = CPM) %>%
  column_to_rownames("transcript_id")

transcript_matrix_log2 <- log2(transcript_matrix + 1)
colnames(transcript_matrix_log2) <- c("Q157R_1", "Q157R_2", "Q157R_3", "WT_1", "WT_2", "WT_3")

annotation_col <- data.frame(Group = c(rep("Q157R", 3), rep("WT", 3)))
rownames(annotation_col) <- colnames(transcript_matrix_log2)

pheatmap(transcript_matrix_log2,
         cluster_rows = TRUE,
         cluster_cols = TRUE,
         scale = "row",
         show_rownames = FALSE,
         annotation_col = annotation_col,
         annotation_colors = list(
           Group = c("Q157R" = "#336B87", "WT" = "#E99787")
         ),
         main = "Dynamic Transcript Expression (WT vs Q157R)")

pdf(file.path(plots_subfolder, "dynamic_transcript_heatmap_WT_vs_Q157R.pdf"), width = 8, height = 10)
print(last_plot())
dev.off()
```
Heat map for filtered genes - top 500
```{r}
# Average transcript CPMs to gene CPMs
gene_matrix <- dynamic_cpm_long %>%
  group_by(associated_gene, sample) %>%
  summarise(mean_CPM = mean(CPM, na.rm = TRUE)) %>%
  ungroup() %>%
  pivot_wider(names_from = sample, values_from = mean_CPM) %>%
  column_to_rownames("associated_gene")

# Log2 transform
gene_matrix_log2 <- log2(gene_matrix + 1)

# Rename sample columns
colnames(gene_matrix_log2) <- c("Q157R_1", "Q157R_2", "Q157R_3", "WT_1", "WT_2", "WT_3")

# Create annotation dataframe
annotation_col_gene <- data.frame(
  Group = c(rep("Q157R", 3), rep("WT", 3))
)
rownames(annotation_col_gene) <- colnames(gene_matrix_log2)

# Make the plot object
p_gene <- pheatmap(gene_matrix_log2,
                   cluster_rows = TRUE,
                   cluster_cols = TRUE,
                   scale = "row",
                   show_rownames = TRUE,
                   annotation_col = annotation_col,
                   annotation_colors = list(
                     Group = c("Q157R" = "#336B87", "WT" = "#E99787")
                     ),
                   main = "Dynamic Gene Expression (WT vs Q157R)")

# Save gene heatmap
pdf(file.path(plots_subfolder, "dynamic_gene_heatmap_WT_vs_Q157R.pdf"), width = 8, height = 10)
print(p_gene)  # important to print the plot inside the pdf
dev.off()

```
Heat map for filtered transcripts - top 50
```{r}
# Select top 50 dynamic transcripts
top_dynamic_transcripts <- dynamic_transcripts %>%
  arrange(PValue) %>%  # You could also use logFC or F if you prefer
  slice_head(n = 50)

# Filter CPM long data to only top 50 transcripts
dynamic_cpm_long_top50 <- dynamic_cpm_long %>%
  filter(transcript_id %in% top_dynamic_transcripts$isoform)

# Reshape back to wide format: transcript_id × sample
transcript_matrix <- dynamic_cpm_long_top50 %>%
  select(transcript_id, sample, CPM) %>%
  pivot_wider(names_from = sample, values_from = CPM) %>%
  column_to_rownames("transcript_id")

# Log2 transform
transcript_matrix_log2 <- log2(transcript_matrix + 1)

# Rename sample columns
colnames(transcript_matrix_log2) <- c("Q157R_1", "Q157R_2", "Q157R_3", "WT_1", "WT_2", "WT_3")

# Create annotation dataframe
annotation_col <- data.frame(
  Group = c(rep("Q157R", 3), rep("WT", 3))
)
rownames(annotation_col) <- colnames(transcript_matrix_log2)

# Make and save the plot
p <- pheatmap(transcript_matrix_log2,
              cluster_rows = TRUE,
              cluster_cols = TRUE,
              scale = "row",
              show_rownames = TRUE,  # Now show rownames since only 50!
              annotation_col = annotation_col,
                   annotation_colors = list(
                     Group = c("Q157R" = "#336B87", "WT" = "#E99787")
                     ),
              main = "Top 50 Dynamic Transcripts (WT vs Q157R)")

pdf(file.path(plots_subfolder, "dynamic_transcript_heatmap_top50_WT_vs_Q157R.pdf"), width = 8, height = 10)
print(p)
dev.off()
```
Heat map for filtered genes - top 50
```{r}
# Average transcript CPMs to gene CPMs
gene_matrix <- dynamic_cpm_long_top50 %>%
  group_by(associated_gene, sample) %>%
  summarise(mean_CPM = mean(CPM, na.rm = TRUE)) %>%
  ungroup() %>%
  pivot_wider(names_from = sample, values_from = mean_CPM) %>%
  column_to_rownames("associated_gene")

# Log2 transform
gene_matrix_log2 <- log2(gene_matrix + 1)

# Rename sample columns
colnames(gene_matrix_log2) <- c("Q157R_1", "Q157R_2", "Q157R_3", "WT_1", "WT_2", "WT_3")

# Create annotation dataframe
annotation_col_gene <- data.frame(
  Group = c(rep("Q157R", 3), rep("WT", 3))
)
rownames(annotation_col_gene) <- colnames(gene_matrix_log2)

# Make and save the plot
p_gene <- pheatmap(gene_matrix_log2,
                   cluster_rows = TRUE,
                   cluster_cols = TRUE,
                   scale = "row",
                   show_rownames = TRUE,
                   annotation_col = annotation_col,
                   annotation_colors = list(
                     Group = c("Q157R" = "#336B87", "WT" = "#E99787")
                     ),
                   main = "Top 50 Dynamic Genes (WT vs Q157R)")

pdf(file.path(plots_subfolder, "dynamic_gene_heatmap_top50_WT_vs_Q157R.pdf"), width = 8, height = 10)
print(p_gene)
dev.off()
```
Create stacked bar plots
```{r}
# Additional Libraries
library(RColorBrewer)

# 1. Only plot genes from non_unique list
genes_to_plot <- unique(non_unique$associated_gene)

# 2. Filter the averaged fractional abundance to these genes
average_frac_abundance_filtered <- average_frac_abundance %>%
  filter(Gene %in% genes_to_plot)

# 3. Define green and brown colors
custom_colors <- c("Q157R" = "#336B87",  # ForestGreen
                   "WT" = "#E99787")     # SaddleBrown

# 4. Loop through genes and plot stacked bar plots
for (gene in genes_to_plot) {
  gene_data <- average_frac_abundance_filtered %>% filter(Gene == gene)
  
  if (nrow(gene_data) == 0) next
  
  p <- ggplot(gene_data, aes(x = group, y = avg_abundance, fill = Transcript)) +
    geom_bar(stat = "identity", position = "stack") +
    labs(title = paste("Stacked Barplot for", gene, "(WT vs Q157R)"),
         x = "Group", y = "Average Fractional Abundance") +
    theme_minimal() +
    guides(fill = guide_legend(reverse = TRUE)) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    scale_y_continuous(expand = c(0, 0)) +
    scale_fill_manual(values = colorRampPalette(c("#336B87", "#E99787"))(length(unique(gene_data$Transcript))))  # green to brown gradient
  
  # Save each plot
  ggsave(file.path(plots_subfolder, paste0(gene, "_stacked_barplot_WT_vs_Q157R.pdf")),
         plot = p, width = 6, height = 5)
}



```

Replicate-level stacked bar plots
```{r}
# 1. Loop through genes again, now for replicate-level plotting
for (gene in genes_to_plot) {
  gene_data_rep <- frac_long %>%
    filter(associated_gene == gene) %>%
    mutate(Sample_Group = paste(group, sample, sep = "_"))
  
  if (nrow(gene_data_rep) == 0) next
  
  # 2. Define colors across isoforms again
  n_transcripts <- length(unique(gene_data_rep$transcript_id))
  mycolors_rep <- colorRampPalette(c("#336B87", "#E99787"))(n_transcripts)  # green to brown gradient
  
  # 3. Plot
  p_rep <- ggplot(gene_data_rep, aes(x = Sample_Group, y = fractional_abundance, fill = transcript_id)) +
    geom_bar(stat = "identity", position = "stack") +
    labs(title = paste("Replicate-level Stacked Barplot for", gene, "(WT vs Q157R)"),
         x = "Sample", 
         y = "Fractional Abundance") +
    theme_minimal() +
    scale_fill_manual(values = mycolors_rep) +
    guides(fill = guide_legend(reverse = TRUE, title = "Transcript")) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    scale_y_continuous(expand = c(0, 0))
  
  # Save replicate plot
  ggsave(file.path(plots_subfolder, paste0(gene, "_replicate_stacked_barplot_WT_vs_Q157R.pdf")),
         plot = p_rep, width = 8, height = 5)
}
```